---
title: "Untitled"
author: "Aqsa Adab"
date: "`r Sys.Date()`"
output: word_document
---
Step1: Exploratery Data Analysis



```{r}
# Load readxl for Excel file reading
library(readxl)
```

Importing the Data

```{r}
# Specify the file path to the Excel file
file_path <- "C:/Users/Aqsa/Downloads/1657875746_day.xlsx"

# Read the Excel file into a data frame
bike_df <- read_excel(file_path)
```


```{r}
# Display the first 5 rows of the bike_df dataset
head(bike_df, 5)
```

Omit the 'casual' and 'registered' variables because 'total_count' already includes the combined count of both.

```{r}
#Create new dataset excluding casual and registered variables
bike_df<-subset(bike_df,select=-c(casual,registered))
head(bike_df,5)
```

Dimension of Dataset
```{r}
#Dimension of Dataset
dim(bike_df)
```
Summary of the Dataset
```{r}
#Summary of the dataset
summary(bike_df)
```

Structure of Dataset
```{r}
#Structure of Dataset
str(bike_df)
```

Rename the columns for better understanding of variables
```{r}
#Rename the columns
names(bike_df)<-c('rec_id','datetime','season','year','month','holiday','weekday','workingday','weather_condition','temp','atemp','humidity','windspeed','total_count')
```


```{r}
#Read the data after renaming the columns.
head(bike_df,5)
```


```{r}
#Typecasting the datetime and numerical attributes to category

bike_df$datetime<- as.Date(bike_df$datetime)
bike_df$year<-as.factor(bike_df$year)
bike_df$month<-as.factor(bike_df$month)
bike_df$season <- as.factor(bike_df$season)
bike_df$holiday<- as.factor(bike_df$holiday)
bike_df$weekday<- as.factor(bike_df$weekday)
bike_df$workingday<- as.factor(bike_df$workingday)
bike_df$weather_condition<- as.factor(bike_df$weather_condition)
```

Missing value analysis
```{r}
#Missing values in dataset
missing_val<-data.frame(apply(bike_df,2,function(x){sum(is.na(x))}))
names(missing_val)[1]='missing_val'
missing_val
```

There are no missing values in the dataset as indicated by the 'missing_val' column. Each variable, including 'rec_id' and 'datetime,' has zero missing values.

```{r}
#ggplot2 is utilized for creating versatile and customizable data visualizations in R, enabling the generation of complex plots with concise and expressive code.
library(ggplot2)
```


```{r}
# Column plot for season-wise monthly distribution of counts
ggplot(bike_df, aes(x = month, y = total_count, fill = season)) +
  theme_bw() +
  geom_col() +
  scale_fill_manual(values = c("#0072B2", "#D55E00", "#009E73", "#CC79A7")) +  # Set custom colors
  labs(x = 'Month', y = 'Total Count', title = 'Season-wise Monthly Distribution of Counts')
```


```{r}
# Column plot for weekday-wise monthly distribution of counts
ggplot(bike_df, aes(x = month, y = total_count, fill = weekday)) +
  theme_bw() +
  geom_col() +
  scale_fill_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854", "#ffd92f", "#e5c494")) +  # Set custom colors
  labs(x = 'Month', y = 'Total Count', title = 'Weekday-wise Monthly Distribution of Counts')
```

Plot Type: Column Chart or Bar Chart

Analysis:

The first plot illustrates the season-wise monthly distribution of bike rental counts. It reveals the variation in counts across different months within each season, with custom colors for better distinction.

The second plot displays the weekday-wise monthly distribution of bike rental counts. It depicts how counts vary across months for each weekday, using custom colors to differentiate weekdays.

We can see from the above plots that there is a seasonal increase in bike rentals during the spring and summer, followed by a decrease during the fall and winter.

Season 1 leads to spring, Season 2 to summer, Season 3 to autumn, and Season 4 to winter in this case.

```{r}
# Violin plot for yearly wise distribution of counts
ggplot(bike_df, aes(x = factor(year), y = total_count, fill = factor(year))) +
  geom_violin() +
  theme_bw() +
  scale_fill_manual(values = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00")) +  # Set custom colors
  labs(x = 'Year', y = 'Total Count', title = 'Yearly Wise Distribution of Counts') +
  theme(legend.position="none", plot.title.position = "plot", plot.title = element_text(hjust = 0.5))
```

Plot Type: Violin Plot

Analysis:
The violin plot illustrates the distribution of bike rental counts over the years. Notably, the distribution is higher in the year 2012 compared to 2011. The custom colors distinguish the violin plots for each year, providing a visual representation of the count distribution.

Holiday wise distribution of counts
```{r}
# Column plot for holiday-wise distribution of counts
ggplot(bike_df, aes(x = factor(holiday), y = total_count, fill = season)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(values = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00")) +  # Set custom colors
  labs(x = 'Holiday', y = 'Total Count', title = 'Holiday Wise Distribution of Counts') +
  theme(legend.position = "none", plot.title.position = "plot", plot.title = element_text(hjust = 0.5))
```

Plot Type: Grouped Column Chart

Analysis:
The bar plot demonstrates the distribution of bike rental counts based on holiday status. Notably, the counts are highest during non-holiday periods (0) compared to holidays (1) across different seasons. Custom colors differentiate the bars representing each season, providing a clear visual distinction in the plot.

```{r}
# Column plot for workingday-wise distribution of counts
ggplot(bike_df, aes(x = factor(workingday), y = total_count, fill = season)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(values = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00")) +  # Set custom colors
  labs(x = 'Workingday', y = 'Total Count', title = 'Workingday Wise Distribution of Counts') +
  theme(legend.position = "none", plot.title.position = "plot", plot.title = element_text(hjust = 0.5))
```

Plot Type: Grouped Column Chart

Analysis:
The bar plot illustrates the distribution of bike rental counts based on working day status. Notably, the counts are relatively higher during working days (1) compared to non-working days (0) across different seasons. Custom colors distinguish the bars representing each season, providing a clear visual representation of the count distribution.

```{r}
# Define custom colors and corresponding labels
custom_colors <- c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00")
custom_labels <- c("Spring", "Summer", "Fall", "Winter")  # Adjust labels as needed

# Column plot for weather_condition-wise distribution of counts
ggplot(bike_df, aes(x = factor(weather_condition), y = total_count, fill = season)) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(values = custom_colors, labels = custom_labels) +  # Set custom colors and labels
  labs(x = 'Weather Condition', y = 'Total Count', title = 'Weather Condition Distribution of Counts') +
  theme(plot.title.position = "plot", plot.title = element_text(hjust = 0.5))
```

Plot Type: Grouped Column Chart

Analysis:
The bar plot visualizes the distribution of bike rental counts based on different weather conditions. Clear and partly cloudy weather contributes to the highest bike rental counts, followed by misty and cloudy weather, and then light snow and light rain conditions. Custom colors and labels are applied to enhance the clarity and interpretation of the plot.

```{r}
# Set a colorful palette
colors <- c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854", "#ffd92f", "#e5c494")

# Create a boxplot with custom colors
boxplot(bike_df$total_count, 
        col = colors,
        main = 'Total_count',
        sub = paste("Outliers: ", boxplot.stats(bike_df$total_count)$out))
```

Plot Type: Boxplot

Analysis:
The boxplot depicts the distribution of the 'total_count' variable, revealing that there are no outliers present. The colorful palette enhances the visual representation of the boxplot, with summary statistics displayed in the subtitle.

```{r}
# Set a colorful palette
colors <- c("#66c2a5", "#fc8d62", "#8da0cb", "#e78ac3", "#a6d854", "#ffd92f", "#e5c494")

# Create a 2x2 layout for box plots
par(mfrow = c(2, 2))

# Box plot for temp outliers
boxplot(bike_df$temp, 
        col = colors[1],  # Use the first color in the palette
        main = "Temperature",
        sub = paste("Outliers: ", boxplot.stats(bike_df$temp)$out),
        at = 1)

# Box plot for humidity outliers
boxplot(bike_df$humidity, 
        col = colors[2],  # Use the second color in the palette
        main = "Humidity",
        sub = paste("Outliers: ", boxplot.stats(bike_df$humidity)$out),
        at = 2)

# Box plot for windspeed outliers
boxplot(bike_df$windspeed, 
        col = colors[3],  # Use the third color in the palette
        main = "Windspeed",
        sub = paste("Outliers: ", boxplot.stats(bike_df$windspeed)$out),
        at = 3)



```

Plot Type: Boxplots (Temperature, Humidity, Windspeed)

Analysis:
The series of boxplots examines outliers in different variables. The boxplot for temperature indicates no outliers. However, in the case of windspeed and humidity, a few outliers are observed, suggesting some data points deviate significantly from the typical range in these variables.

Replace and imputate the outliers
```{r}
#load the DMwR library
library(DMwR)
#create subset for windspeed and humidity variable
wind_hum<-subset(bike_df,select=c('windspeed','humidity'))
#column names of wind_hum
cnames<-colnames(wind_hum)
for(i in cnames){
    val=wind_hum[,i][wind_hum[,i] %in% boxplot.stats(wind_hum[,i])$out] #outlier values
    wind_hum[,i][wind_hum[,i] %in% val]= NA  # Replace outliers with NA 
}
#Imputating the missing values using mean imputation method
wind_hum$windspeed[is.na(wind_hum$windspeed)]<-mean(wind_hum$windspeed,na.rm=T) 
wind_hum$humidity[is.na(wind_hum$humidity)]<-mean(wind_hum$humidity,na.rm=T)
```
Combaining the imputated dataset and original dataset
```{r}
#Remove the windspeed and humidity variable in order to replace imputated data
new_df<-subset(bike_df,select=-c(windspeed,humidity))
#Combined new_df and wind_hum data frames
bike_df<-cbind(new_df,wind_hum)
head(bike_df,5)

```
Normal probability plot

Normal probability plot is a graphical technique to identify substantive departures from normality and also it tells about goodness of fit.

```{r}
#Quintle-Quintle normal plot
qqnorm(bike_df$total_count)
#Quintle-Quintle line
qqline(bike_df$total_count)
```
From the above probability plot, we can observed that some target variable data points are deviates from normality.


Correlation Matrix

Correlation matrix is tells about linear relationship between attributes and help us to build better models.
```{r}
#load the corrgram for correlation
library(corrgram)
#Correlation plot
corrgram(bike_df[,10:14],order=F,upper.panel=panel.pie,text.panel=panel.txt,main='Correlation Plot')
```


Plot Type: Correlation Plot

Analysis:
The correlation plot visually represents the correlation between different features in the dataset. It reveals both positive and negative correlations among the variables. Specifically, 'temp' and 'atemp' are highly positively correlated, indicating redundancy in information. Consequently, the decision is made to exclude the 'atemp' variable from further analysis.



Split the dataset into train and test dataset

```{r}
#load the purrr library for functions and vectors
library(purrr)
#Split the dataset based on simple random resampling
train_index<-sample(1:nrow(bike_df),0.7*nrow(bike_df))
train_data<-bike_df[train_index,]
test_data<-bike_df[-train_index,]
dim(train_data)
dim(test_data)
```
train and test data

```{r}
#Read the train and test data
head(train_data,5)

```


```{r}
head(test_data,5)
```
Selecting the required independent and dependent variables

```{r}
#Create a new subset for train attributes 
train<-subset(train_data,select=c('season','year','month','holiday', 'weekday','workingday','weather_condition','temp','humidity','windspeed','total_count'))
#Create a new subset for test attributes
test<-subset(test_data,select=c('season','year','month','holiday','weekday','workingday','weather_condition','temp','humidity','windspeed','total_count'))
head(train,5)
head(test,5)
```

Train and test categorical and numerical attributes

```{r}
#create a new subset for train categorical attributes
train_cat_attributes<-subset(train,select=c('season','holiday','workingday','weather_condition','year'))
#create a new subset for test categorical attributes
test_cat_attributes<-subset(test,select=c('season','holiday','workingday','weather_condition','year'))
#create a new subset for train numerical attributes
train_num_attributes<-subset(train,select=c('weekday','month','temp','humidity','windspeed','total_count'))
#create a new subset for test numerical attributes
test_num_attributes<-subset(test,select=c('weekday','month','temp', 'humidity','windspeed','total_count'))
```

Encoding the categorical features

Train_encoded_attributes

```{r}
#load the caret library
library(caret)
#other variables along with target variable to get dummy variables
othervars<-c('month','weekday','temp','humidity','windspeed','total_count')
set.seed(2626)
#Categorical variables
vars<-setdiff(colnames(train),c(train$total_count,othervars))
#formula pass through encoder to get dummy variables
f <- paste('~', paste(vars, collapse = ' + '))
#encoder is encoded the categorical variables to numeric
encoder<-dummyVars(as.formula(f), train)
#Predicting the encode attributes
encode_attributes<-predict(encoder,train)
#Binding the train_num_attributes and encode_attributes
train_encoded_attributes<-cbind(train_num_attributes,encode_attributes)
head(train_encoded_attributes,5)
```

Test_encoded_attributes
```{r}
set.seed(5662)
#Categorical variables
vars<-setdiff(colnames(test),c(test$total_count,othervars))
#formula pass through encoder to get dummy variables
f<- paste('~',paste(vars,collapse='+'))
#Encoder is encoded the categorical variables to numeric
encoder<-dummyVars(as.formula(f),test)
#Predicting the encoder attributes
encode_attributes<-predict(encoder,test)
#Binding the test_num_attributes and encode_attributes
test_encoded_attributes<-cbind(test_num_attributes,encode_attributes)
head(test_encoded_attributes,5)
```

Modelling the training dataset

Random Forest
```{r}
set.seed(6788271)
#load the randomForest library
library(randomForest)
#training the model
rf_model<-randomForest(total_count~.,train_encoded_attributes,importance=TRUE,ntree=200)
rf_model
```
Cross validation prediction for Random Forest

```{r}
options(warn=-1)
set.seed(6772)
library(randomForest)
#load the ranger library for random forest CV
library(ranger)
#Cross validation resampling method
train.control<-trainControl(method='CV',number=3)
#Cross validation prediction
rf_CV_predict<-train(total_count~.,train_encoded_attributes,method='ranger',trControl=train.control)
rf_CV_predict

```

Cross Validation prediction plot
```{r}
#Cross validation prediction plot
residuals<-resid(rf_CV_predict)
plot(y_train,residuals,xlab='Observed',ylab='Residuals',main='Cross validation prediction plot')
abline(0,0)
```
Cross validation prediction plot tells about finite variance between actual target value and predicted target value. In this plot, some data points are have same finite variance between them and for some are not have it.

Model performance on the test dataset
```{r}
set.seed(7889)
#Predicting the model
rf_predict<-predict(rf_model,test_encoded_attributes[,-c(6)])
head(rf_predict,5)
```
Root mean squared error and mean absolute error

```{r}
set.seed(667)
#Root mean squared error
rmse<-RMSE(y_test,rf_predict)
print(rmse)
mae<-MAE(y_test,rf_predict)
print(mae)
```
Residual plot

```{r}
#Residual plot
residuals<-y_test-rf_predict
plot(y_test,residuals,xlab='Observed',ylab='Residuals',main='Residual plot')
abline(0,0)
```
Residual plot tells about finite variance between actual target value and predicted target value. In this plot, some data points are have same finite variance between them and for some are not have it.

Final model for predicting the bike rental count on daily basis

When we compare the root mean squared error and mean absolute error of all 3 models, the random forest model has less root mean squared error and mean absolute error. So, finally random forest model is best for predicting the bike rental count on daily basis.

```{r}
Bike_predictions=data.frame(y_test,rf_predict)
write.csv(Bike_predictions,'Bike_Renting_R.CSV',row.names=F)
Bike_predictions
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```
